HO unification from object language to meta language

### 2024-03-28 paper/main.tex
update tex

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\section{Enforcing Invariant~\ref{invariant:arity}}
\label{sec:invariant1}

@@

We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, we can remark that
this compilation breaks \cref{inv:map-bijective}, since

Deduplicate mapping code etc...

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\section{Enforcing Invariant~\ref{inv:uvaarity}}
\label{sec:invariant1}

@@

We see that the \elpiIn{maybe-eta} as detected $\lambda xy.F \appsep y \appsep
x$ and $\lambda x.f\appsep (F \appsep x) \appsep x$ and replaced them with
respectively the \Ho vars $X$ and $Z$. $X$ is linked with $\lambda x.Y_x$, $Y$
has arity $1$ and is $\eta$-linked with $\lambda y.H\appsep y \appsep x$ and $Z$
is linked to the term $\lambda x.f \appsep G_{x}~x$. However, the mapping
returned by the compilation, does not breaks \cref{inv:map-bijective}: the \Fo{}
variable $F$ is mapped to two different \Ho variables. To address this problem
and enforce \cref{inv:map-bijective}, we clean the mapping with a second phase
after the compilation. This phase is called \elpiIn{map-deduplication}.

\newcommand{\cv}{\ensuremath{\mathrm{cv}}}
\newcommand{\ev}{\ensuremath{\mathrm{ev}}}
\newcommand{\ar}{\ensuremath{\mathrm{ar}}}

Before formally defining this procedure, we need to define some auxiliary
relations. Let \mapStore be the list of mapping and let $m \in \mapStore$, such
that $m = \mapping{A}{\textbf{A}}{n}$, we define $\cv, \ev, ar$ such that
$\cv(m) = A$, $\ev(m) = \textbf{A}$ and $\ar(m) = n$. Let $\langle m_1, m_2
\rangle \in \mapStore$ such that $\ar(m_1) < \ar(m_2)$ and $n = ar(m_1) -
ar(m_2) - 1$. We also let $A^i$ be a fresh \Ho variable. We define the
\elpiIn{make-eta-link} relation taking two mappings $\langle m_1, m_2 \rangle$
and returning  \linketa such that:

$$ 
  \forall i \in [0 \dots n], \left\{ 
    \begin{array}{ll}
      \linketaM{\phantom{\Gamma:x_0\dots x_{i-1}}}
              {\ev(m_1)}
              {\lambda x.A^0_x} & if~ i = 0 \\
      \linketaM{\Gamma:x_0\dots x_{i-1}}
               {A^i_{\Gamma}}
               {\lambda x.A^{i+1}_{\Gamma\cup x}} & if~ 0 < i < n\\
      \linketaM{\Gamma:x_0\dots x_{i}\phantom{_{-1}}}
               {A^{i-1}_{\Gamma}}
               {\lambda x.\ev(m_2)_{\Gamma\cup x}} & if~ i = n\\
    \end{array} 
  \right. 
$$

More concretely, we are saying that for any two mappings, we build as many
\linketa as the difference of the arities between the two mappings. This links
are constructed in such a way that the \Ho variable $v$ with lowest arity is
linked to a fresh variable eta-expended variable $A^0$ having the scope of $v$.
This variable $A^0$ is then linked to a fresh variable $A^{1}$ with same scope
of $A^0$ and so on. The last link is built between the $A^{n-1}$ (where $n$ is
the difference of arities between the two mappings) and the \Ho variable $u$
with higher arity in the two mappings being considered.

\begin{definition}[map-deduplication]
  Forall mappings $\langle m_1, m_2 \rangle \in \mapStore$, such that $\cv(m_1) =
  \cv(m_2)$, the list of \linketa $L$ is created thanks to
  \elpiIn{make-eta-link}$m_1~m_2~L$ and is added to \linkStore. Then $m_1$ is
  removed from \mapStore.
\end{definition}

If we take back the example give at the beginning of this section, we can
deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}.

The complete problem to run for resolution is now:

\begin{description}
  \item[Unif problems:] $X = \lambda x.\lambda y.x, Z = f\appsep(\lambda x.x)$
  \item[Mappings:] $\mapping{F}{H}{2}$  
  \item[Links:] $l_1 = \linketaM{x}{Y_{x}}{\lambda y.H_{y~x}}$,\\
                $l_2 = \linketaM{}{X}{\lambda x.Y_{x}}$,\\
                $l_3 = \linketaM{}{Z}{\lambda x.f \appsep G_{x}~x}$\\
                $l_4 = \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}$
\end{description}

After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
This assignment makes $l_2$ to progress since the \lhs is materialized and by
unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
\elpiIn{progress} fixpoint terminates. Next, the second unification problem is
run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
$Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
As last step, we can note that $G$

TODO: dire che preserviamo l'invariante che tutte le variable sono fully-applied
